Manuscript 1 code

### 2024-10-25 manuscript_1_code.Rmd
cleaned code up through figure 3 with new functions for extracting ranges of data (see functions.R)

diversity_ranges_quad <- extract_ranges(alpha_diversity_quad_macro, "habitat", c("richness", "cover_trans"))

### Setup
```{r read and prep functional group data}
algae_fg_raw2 <- read_csv(here("data","algae_functional_groupings_092724.csv"))

@@

#richness at every quad/year combination 
fg_summary_wide$fg_richness_quad <- vegan::specnumber(fg_summary_wide[5:14])

unique(fg_summary_wide$fg_richness_quad)
#0 to 4

@@

#richness at every site/year combination ---> NOT AVERAGED,
fg_summary_site_wide_SUM$fg_richness_site <- vegan::specnumber(fg_summary_site_wide_SUM[4:13])

fg_summary_site_wide <- fg_summary_site_wide_SUM

#merge Noam's data with functional group summary 
functional_richness_quad_cover <- merge(alpha_diversity_quad_macro, fg_summary_wide,
                                        by = c("location", "year", "site", "habitat"))

#merge site level cover with site functional group richness
functional_richness_site_cover<- merge(alpha_diversity_site_macro, fg_summary_site_wide,
                                by = c("year", "site",  "habitat"))
```

@@

cover_mod_fg <- glmmTMB(cover_trans ~ fg_richness_quad*habitat + (1|site/location) + (1|year), family = beta_family(), data = functional_richness_quad_cover)

@@

r.squaredGLMM(cover_mod_fg) # 0.64, 0.65
emtrends(cover_mod_fg, pairwise ~ habitat, var = "fg_richness_quad") 
# contrast                    estimate     SE  df z.ratio p.value
# Fringing - Backreef            0.214 0.0266 Inf   8.051  <.0001
# Fringing - Forereef 10m        0.346 0.0261 Inf  13.252  <.0001
# Fringing - Forereef 17m        0.530 0.0262 Inf  20.244  <.0001
# Backreef - Forereef 10m        0.132 0.0231 Inf   5.708  <.0001
# Backreef - Forereef 17m        0.316 0.0232 Inf  13.617  <.0001
# Forereef 10m - Forereef 17m    0.184 0.0223 Inf   8.241  <.0001

@@

## figure 3b plot setup
```{r figure 3b setup}
diversity_ranges_quad_fg <- functional_richness_quad_cover %>%
  group_by(habitat) %>%
  summarise(min_fg_richness_quad = min(fg_richness_quad),
            max_fg_richness_quad = max(fg_richness_quad),
            min_cover= min(cover_trans),
            max_cover = max(cover_trans))

cover_emm_fg <- emmip(cover_mod_fg,
                        habitat ~ fg_richness_quad,
                        at = list(fg_richness_quad = seq(0,10,1)), plotit = F, CIs = T)

filtered_cover_effects_fg <- cover_emm_fg %>%
  left_join(diversity_ranges_quad_fg, by = c("habitat")) %>%
  filter(fg_richness_quad >= min_fg_richness_quad & fg_richness_quad <= max_fg_richness_quad)

@@

```{r figure 3b plot}
(figure_3b <- 
  ggplot() +
  geom_point(data = functional_richness_quad_cover,
             aes(x = fg_richness_quad, y = cover_trans, colour = habitat), size = 1, alpha = 0.005) + 
  geom_line(data = filtered_cover_effects_fg, aes(x = fg_richness_quad, y = inv_logit(yvar), colour = habitat)) +
  geom_ribbon(data = filtered_cover_effects_fg, aes(x = fg_richness_quad, ymin = inv_logit(LCL), ymax = inv_logit(UCL), fill = habitat),
              alpha = 0.3) +
  scale_colour_manual(values = habitat_colours) +
  scale_fill_manual(values = habitat_colours) +
  labs(y = "", x = "", title = "") +
  trends_theme
         )

@@

diversity_stability <- alpha_diversity_quad_macro %>%
  dplyr::group_by(location, habitat, site) %>%
  dplyr::summarise(
            richness_mean = mean(richness), 
            shannon_mean = mean(shannon), 
            cover_mean = mean(macroalgal_prop_cover),
            evenness_mean = mean(evenness, na.rm = T),
            cover_cv = CV(macroalgal_prop_cover),
            cover_stability = 1/CV(macroalgal_prop_cover))

@@

dss_ranges<- diversity_stability_synchrony %>%
  group_by(habitat) %>%
  summarise(min_richness = min(richness_mean),
            max_richness = max(richness_mean),
            min_evenness = min(richness_mean, na.rm = TRUE),
            max_evenness = max(richness_mean, na.rm = TRUE),
            min_stability= min(cover_stability, na.rm = TRUE),
            max_stability = max(cover_stability, na.rm = TRUE),
            min_synchrony = min(synchrony_trans, na.rm = TRUE),
            max_synchrony = max(synchrony_trans, na.rm = TRUE))
```

@@

filtered_synch_cov_effects <- synch_cov_emm %>%
  left_join(dss_ranges, by = c("habitat")) %>%
  filter(synchrony >= min_synchrony & synchrony <= max_synchrony)
```

@@

### Synchrony setup site level
```{r calculate syncrony site_level}
# make er long
macro_long_data_site <- pivot_longer(
  data = site_macro,
  cols = grep("acanthophora_spicifera", colnames(site_macro)):grep("valonia_ventricosa", colnames(site_macro)),
  names_to = "taxa",
  values_to = "percent_cover") %>% 
  mutate(prop_cover = percent_cover/100) %>% 
  select(-percent_cover)

@@

diversity_stability_site <- alpha_diversity_site_macro %>%
  dplyr::group_by(site_habitat, habitat, site) %>%
  dplyr::summarise(
            richness_mean = mean(richness), 
            shannon_mean = mean(shannon), 
            cover_mean = mean(prop_cover),
            evenness_mean = mean(evenness),
            cover_cv = CV(prop_cover),
            cover_stability = 1/CV(prop_cover))

@@

dss_ranges_site <- diversity_stability_synchrony_site %>%
  group_by(habitat) %>%
  summarise(min_richness = min(richness_mean),
            max_richness = max(richness_mean),
            min_evenness = min(evenness_mean, na.rm = T),
            max_evenness = max(evenness_mean, na.rm = T),
            min_stability= min(cover_stability),
            max_stability = max(cover_stability),
            min_synchrony = min(synchrony_trans),
            max_synchrony = max(synchrony_trans))
```

@@

### setup
```{r site richness cover setup}
diversity_ranges_site <- alpha_diversity_site_macro %>%
  group_by(habitat) %>%
  summarise(min_richness = min(richness),
            max_richness = max(richness),
            min_cover= min(cover_trans),
            max_cover = max(cover_trans))

@@

filtered_site_cover_effects <- site_cover_emm %>%
  left_join(diversity_ranges_site, by = c("habitat")) %>%
  filter(richness >= min_richness & richness <= max_richness)

@@

site_cover_mod_fg <- glmmTMB(cover_trans ~ fg_richness_site*habitat + (1|site_habitat) + (1|year), family = beta_family(), data = functional_richness_site_cover)

@@

hist(residuals(site_cover_mod_fg)) # positive skew
plot(residuals(site_cover_mod_fg) ~ fitted(site_cover_mod_fg)) # a little heteroscedastic but can't be less than 0
r.squaredGLMM(site_cover_mod_fg) # 0.180, 0.563
emtrends(site_cover_mod_fg, pairwise ~ habitat, var = "fg_richness_site") # no difference between the habitats
# Fringing - Backreef           0.0379 0.0825 Inf   0.459  0.9679
# Fringing - Forereef 10m       0.1846 0.0930 Inf   1.985  0.1935
# Fringing - Forereef 17m       0.0933 0.0960 Inf   0.972  0.7655
# Backreef - Forereef 10m       0.1467 0.0986 Inf   1.488  0.4447
# Backreef - Forereef 17m       0.0554 0.1018 Inf   0.545  0.9480
# Forereef 10m - Forereef 17m  -0.0913 0.1093 Inf  -0.835  0.8377

```
```{r site fg richness cover plot setup}
fg_ranges_site <- functional_richness_site_cover %>%
  group_by(habitat) %>%
  summarise(min_fg_richness_site = min(fg_richness_site),
            max_fg_richness_site = max(fg_richness_site),
            min_cover= min(cover_trans),
            max_cover = max(cover_trans))

site_cover_fg_emm <- emmip(site_cover_mod_fg,
                        habitat ~ fg_richness_site,
                        at = list(fg_richness_site = seq(0,17,1)), plotit = F, CIs = T)

filtered_site_fg_cover_effects <- site_cover_fg_emm %>%
  left_join(fg_ranges_site, by = c("habitat")) %>%
  filter(fg_richness_site >= min_fg_richness_site & fg_richness_site <= max_fg_richness_site)
```

@@

```{r figure 3e plot }
(figure_3e <- 
  ggplot() +
  geom_point(data = functional_richness_site_cover,
             aes(x = fg_richness_site, y = cover_trans, colour = habitat), size = 1, alpha = 0.3) + 
  geom_line(data = filtered_site_fg_cover_effects, aes(x = fg_richness_site, y = inv_logit(yvar), colour = habitat),
            size = 1.5) +
  geom_ribbon(data = filtered_site_fg_cover_effects, aes(x = fg_richness_site, ymin = inv_logit(LCL), ymax = inv_logit(UCL), fill = habitat),
              alpha = 0.3) +
  scale_colour_manual(values = habitat_colours) +
  scale_fill_manual(values = habitat_colours) +
  labs(y = "", x = "Functional Richness", title = "") +
  trends_theme
         )
#ggsave(filename = "output/figure_3e_v2.png", figure_3e, height = 10, width = 14)
```

@@

filtered_synch_cov_effects_site <- synch_cov_emm_site %>%
  left_join(dss_ranges_site, by = c("habitat")) %>%
  filter(synchrony >= min_synchrony & synchrony <= max_synchrony)
```

>>>>

```{r read and prep functional group data}
algae_fg_raw2 <- read_csv(here("data","algae_functional_groupings_092724.csv"))

@@

#richness at every quad/year combination 
fg_summary_wide$functional_richness <- vegan::specnumber(fg_summary_wide[5:14])

unique(fg_summary_wide$functional_richness)
#0 to 4

@@

#richness at every site/year combination ---> NOT AVERAGED,
fg_summary_site_wide_SUM$functional_richness <- vegan::specnumber(fg_summary_site_wide_SUM[4:13])

#merge Noam's data with functional group summary 
alpha_diversity_quad_macro <- merge(alpha_diversity_quad_macro,
                                    fg_summary_wide[,c("location", "year", "site",  "habitat", "functional_richness")],
                                    by = c("location", "year", "site", "habitat"))

#merge site level cover with site functional group richness
alpha_diversity_site_macro<-merge(alpha_diversity_site_macro,
                                  fg_summary_site_wide_SUM[,c("year", "site",  "habitat", "functional_richness")],
                                  by = c("year", "site",  "habitat"))
```

diversity_ranges_quad <- extract_ranges(alpha_diversity_quad_macro, "habitat", c("richness", "cover_trans", "functional_richness"))

### Setup

@@

cover_mod_fg <- glmmTMB(cover_trans ~ functional_richness*habitat + (1|site/location) + (1|year), family = beta_family(), data = alpha_diversity_quad_macro)

@@

r.squaredGLMM(cover_mod_fg) # 0.64, 0.65
emtrends(cover_mod_fg, pairwise ~ habitat, var = "functional_richness") 
# contrast                    estimate     SE  df z.ratio p.value
# Fringing - Backreef            0.214 0.0266 Inf   8.051  <.0001
# Fringing - Forereef 10m        0.346 0.0261 Inf  13.252  <.0001
# Fringing - Forereef 17m        0.530 0.0262 Inf  20.244  <.0001
# Backreef - Forereef 10m        0.132 0.0231 Inf   5.708  <.0001
# Backreef - Forereef 17m        0.316 0.0232 Inf  13.617  <.0001
# Forereef 10m - Forereef 17m    0.184 0.0223 Inf   8.241  <.0001

@@

## figure 3b plot setup
```{r figure 3b setup}

cover_emm_fg <- emmip(cover_mod_fg,
                        habitat ~ functional_richness,
                        at = list(functional_richness = seq(0,10,1)), plotit = F, CIs = T)

filtered_cover_effects_fg <- filter_ranges(cover_emm_fg, diversity_ranges_quad, "habitat", "functional_richness")

@@

```{r figure 3b plot}
(figure_3b <- 
  ggplot() +
  geom_point(data = alpha_diversity_quad_macro,
             aes(x = functional_richness, y = cover_trans, colour = habitat), size = 1, alpha = 0.005) + 
  geom_line(data = filtered_cover_effects_fg, aes(x = functional_richness, y = inv_logit(yvar), colour = habitat)) +
  geom_ribbon(data = filtered_cover_effects_fg, aes(x = functional_richness, ymin = inv_logit(LCL), ymax = inv_logit(UCL), fill = habitat),
              alpha = 0.3) +
  scale_colour_manual(values = habitat_colours) +
  scale_fill_manual(values = habitat_colours) +
  labs(y = "", x = "", title = "") +
  trends_theme
         )

@@

diversity_stability <- alpha_diversity_quad_macro %>%
  dplyr::group_by(location, habitat, site) %>%
  dplyr::summarise(
            richness_mean = mean(richness), 
            shannon_mean = mean(shannon), 
            cover_mean = mean(macroalgal_prop_cover),
            functional_richness_mean = mean(functional_richness),
            cover_cv = CV(macroalgal_prop_cover),
            cover_stability = 1/CV(macroalgal_prop_cover))

@@

dss_ranges <- extract_ranges(diversity_stability_synchrony, "habitat",
                             c("richness_mean", "functional_richness_mean", "cover_stability", "synchrony"))

@@

filtered_synch_cov_effects <- filter_ranges(synch_cov_emm, dss_ranges, "habitat", "synchrony")

@@

### Synchrony setup site level
```{r calculate syncrony site_level}
# pivot longer
macro_long_data_site <- pivot_longer(
  data = site_macro,
  cols = grep("acanthophora_spicifera", colnames(site_macro)):grep("valonia_ventricosa", colnames(site_macro)),
  names_to = "taxa",
  values_to = "percent_cover") %>% 
  mutate(prop_cover = percent_cover/100) %>% 
  select(-percent_cover)

@@

diversity_stability_site <- alpha_diversity_site_macro %>%
  dplyr::group_by(site_habitat, habitat, site) %>%
  dplyr::summarise(
            richness_mean = mean(richness), 
            shannon_mean = mean(shannon), 
            cover_mean = mean(prop_cover),
            cover_cv = CV(prop_cover),
            cover_stability = 1/CV(prop_cover))

@@

dss_ranges_site <- extract_ranges(diversity_stability_synchrony_site, "habitat",
                                  c("richness_mean", "cover_stability", "synchrony"))

@@

### setup
```{r site richness cover setup}
diversity_ranges_site <- extract_ranges(alpha_diversity_site_macro, "habitat", c("richness", "functional_richness", "cover_trans"))

@@

filtered_site_cover_effects <- filter_ranges(site_cover_emm, diversity_ranges_site, "habitat", "richness")

@@

site_cover_mod_fg <- glmmTMB(cover_trans ~ functional_richness*habitat + (1|site_habitat) + (1|year), family = beta_family(), data = alpha_diversity_site_macro)

@@

hist(residuals(site_cover_mod_fg)) # positive skew
plot(residuals(site_cover_mod_fg) ~ fitted(site_cover_mod_fg)) # a little heteroscedastic but can't be less than 0
r.squaredGLMM(site_cover_mod_fg) # 0.180, 0.563
emtrends(site_cover_mod_fg, pairwise ~ habitat, var = "functional_richness") # no difference between the habitats
# Fringing - Backreef           0.0379 0.0825 Inf   0.459  0.9679
# Fringing - Forereef 10m       0.1846 0.0930 Inf   1.985  0.1935
# Fringing - Forereef 17m       0.0933 0.0960 Inf   0.972  0.7655
# Backreef - Forereef 10m       0.1467 0.0986 Inf   1.488  0.4447
# Backreef - Forereef 17m       0.0554 0.1018 Inf   0.545  0.9480
# Forereef 10m - Forereef 17m  -0.0913 0.1093 Inf  -0.835  0.8377

```
```{r site fg richness cover plot setup}

site_cover_fg_emm <- emmip(site_cover_mod_fg,
                        habitat ~ functional_richness,
                        at = list(functional_richness = seq(0,17,1)), plotit = F, CIs = T)

filtered_site_fg_cover_effects <- filter_ranges(site_cover_fg_emm, diversity_ranges_site, "habitat", "functional_richness")

@@

```{r figure 3e plot }
(figure_3e <- 
  ggplot() +
  geom_point(data = alpha_diversity_site_macro,
             aes(x = functional_richness, y = cover_trans, colour = habitat), size = 1, alpha = 0.3) + 
  geom_line(data = filtered_site_fg_cover_effects, aes(x = functional_richness, y = inv_logit(yvar), colour = habitat),
            size = 1.5) +
  geom_ribbon(data = filtered_site_fg_cover_effects, aes(x = functional_richness, ymin = inv_logit(LCL), ymax = inv_logit(UCL), fill = habitat),
              alpha = 0.3) +
  scale_colour_manual(values = habitat_colours) +
  scale_fill_manual(values = habitat_colours) +
  labs(y = "", x = "Functional Richness", title = "") +
  trends_theme
         )
#ggsave(filename = "output/figure_3e_v2.png", figure_3e, height = 10, width = 14)
```

@@

filtered_synch_cov_effects_site <- filter_ranges(synch_cov_emm_site, dss_ranges_site, "habitat", "synchrony")
```
